Characterizing Signal Loss, Error, and Bias in the 21cm Reionization \\Power Spectrum: A Revised Study of PAPER-64

### 2018-06-07 PAPER.tex
Made some changes to OQE descriptions

The optimal quadratic estimator prescription is then to compute
\begin{equation}
\label{eq:OQE}
\widehat{p}_{\alpha}  =  ({\F^{-1}})_\alpha^{\beta} (\widehat{q}_{\beta} - \widehat{b}_{\beta} )
\end{equation}
where $\F$ is the Fisher matrix 
\begin{equation}
F^{\alpha \beta} \equiv \frac{1}{2} \textrm{tr} \left( \C^{-1} \frac{\partial\textbf{C}}{\partial p_{\alpha}} \C^{-1} \frac{\partial\textbf{C}}{\partial p_{\beta}} \right)
\end{equation}
and
\begin{equation}
\label{eq:OQEQuadratic}
\hat{q}_\alpha =  \half \textbf{x}^\dagger \invC \PDeriv{\C}{p^\alpha}  \invC \textbf{x} 
\end{equation}
and
\begin{equation}
\label{eq:OQELinear}
\widehat{b}_{\alpha} = \half \Tr\left[ (\C - \mathbf{S}) \invC \PDeriv{\C}{p^\alpha} \invC \right].
\end{equation}
This estimator is the minimum variance (smallest error bar) estimate of the power spectrum  subject to the constraint that it is also unbiased; that is, the ensemble average of the estimator is equal to its true value
\begin{equation}
\langle \widehat{p}_{\alpha} \rangle = p_\alpha.
\end{equation}
\citep{tegmark_et_al1997a,bond_et_al1998}

It is clear that in order to obtain an unbiased estimate in the presence of contaminants, the estimator defined by Equation \eqref{eq:OQE} must be capable of ``suppressing'' or ``removing'' their effects. The way in which the optimal estimator can suppress contaminates is illustrated with a toy model in Appendix \ref{sec:icw_appendix}. In this appendix, we show that the effect of the weighting in Equation \ref{eq:OQEQuadratic} is to project out the modes of $\textbf{U}$ with a different covariance structure than $\mathbf{S}$ in the power spectrum estimate, and the effect of Equation \ref{eq:OQELinear} is to subtract out the remaining bias. Similar effects for a more realistic model of the EoR and foregrounds are shown in \citet{liu_tegmark2011}.

@@

%We do this when forming the un-normalized power spectrum estimate $\widehat{q}_{\alpha}$:
Motivated by the desire to retain the desirable behavior of suppressing contributions of $\mathbf{U}$ to estimates of the EoR power spectrum, we note that is possible to define a more generalized version of the quadratic estimator 
%which may still produce an unbiased estimate, 
where Equation \eqref{eq:OQEQuadratic} is replaced by
\begin{equation}
\label{eq:qhat}
\widehat{q}_{\alpha} = \frac{1}{2}\textbf{x}^{\dagger}\textbf{R}\textbf{Q}^{\alpha}\textbf{R}\textbf{x}.
\end{equation}
where $\textbf{R}$ is a weighting matrix chosen by the data analyst.  For example, inverse covariance weighting (the optimal form of QE) would set $\textbf{R} \equiv \textbf{C}^{-1}$ and a uniform-weighted case would use $\textbf{R} \equiv \textbf{I}$, the identity matrix.
The matrix $\textbf{Q}^{\alpha}$ encodes the dependence of the covariance on the power spectrum parameters $\partial \C/\partial p_\alpha$ but in practice also does other things, including implementing a transform of the frequency-domain visibilities to $\mathbf{k}$-space, taking into account cosmological scalings, and converting the visibilities from Jy to Kelvin.
%with dimensions ($N_{f}$,$N_{f}$) --- as

%%% COME BACK HERE
% Need to still explain the motivation for this
With an appropriate normalization 
%We normalize our power spectrum estimates using the matrix $\textbf{M}$:
\begin{equation}
\label{eq:phat}
\widehat{\textbf{P}} = \textbf{M}\widehat{\textbf{q}},
\end{equation}
$\widehat{\textbf{P}}$ is the estimate of the true power spectrum $\textbf{P}$. The data analyst has a choice for $
\textbf{M}$. For simplicity in this section we choose $\textbf{M}$ to be diagonal, although we explore other cases for the analysis 
of PAPER-64 data as explained in Section \ref{sec:Bias}, subject to the constraint that each element of $\widehat{\textbf{P}}$ is a weighted sum of the elements of $\textbf{P}$ with weights that sum to unity.
% JEA - is one really free to choose M and R independently?

@@

For the case of a mode which can be safely assumed to be predominantly a foreground, its presence in the covariance matrix will result in the desired suppression via a kind of projection (c.f. Equations \ref{eq:IdealToyModelCovariance} and \ref{eq:IdealToyModelEstimator});  whether or not it is strongly correlated with the the actual data vector is irrelevant.  However, in the case the covariance matrix is empirically estimated, the eigenmodes of $\textbf{C}_{\rm EoR}$ will both be incorrect and correlated with the data.  Incorrect eigenmodes of $\textbf{C}_{\rm EoR}$ will lead to non-minimum variance estimates, but will not produce the suppression of the power spectrum amplitude seen.  As indicated in Appendix \ref{sec:sigloss_appendix}, however, having $\widehat{\mathbf{C}}$ correlated with the data vector $\mathbf{x}$ also produces a kind of projection of power in the {\it non}-foreground modes from the resulting power spectrum estimate, thus producing an estimate that is biased low.  In short, {\it if the covariance is computed from the data itself, it carries the risk of over-fitting information in the data and introducing a multiplicative bias (per $k$) to estimates of the signal.}

>>>>

The optimal quadratic estimator prescription is then to compute
\begin{equation}
\label{eq:OQE}
\widehat{p}_{\alpha}  =  ({\F^{-1}})_\alpha^{\beta} (\widehat{q}_{\beta} - \widehat{b}_{\beta} )
\end{equation}
where $\F$ is the Fisher matrix 
\begin{equation}
F^{\alpha \beta} \equiv \frac{1}{2} \textrm{tr} \left( \C^{-1} \frac{\partial\textbf{C}}{\partial p_{\alpha}} \C^{-1} \frac{\partial\textbf{C}}{\partial p_{\beta}} \right)
\end{equation}
and
\begin{equation}
\label{eq:OQEQuadratic}
\hat{q}_\alpha =  \half \textbf{x}^\dagger \invC \PDeriv{\C}{p^\alpha}  \invC \textbf{x} 
\end{equation}
and
\begin{equation}
\label{eq:OQELinear}
\widehat{b}_{\alpha} = \half \Tr\left( \mathbf{U} \invC \PDeriv{\C}{p^\alpha} \invC \right).
\end{equation}
This estimator is the minimum variance (smallest error bar) estimate of the power spectrum  subject to the constraint that it is also unbiased; that is, the ensemble average of the estimator is equal to its true value
\begin{equation}
\label{eq:super_unbiased}
\langle \widehat{p}_{\alpha} \rangle = p_\alpha.
\end{equation}
\citep{tegmark_et_al1997a,bond_et_al1998}

It is clear that in order to obtain an unbiased estimate in the presence of contaminants, the estimator defined by Equation \eqref{eq:OQE} must be capable of ``suppressing'' or ``removing'' their effects. By construction, the subtraction of the residual foreground and noise bias accomplishes this, removing any additive bias. However, the $\mathbf{C}^{-1}$ piece of Equation \eqref{eq:OQEQuadratic} also has the effect of suppressing residual foregrounds and noise, in both the additive bias and any contributions the residuals may have to the variance. The way in which the optimal estimator accomplishes this is illustrated with a toy model in Appendix \ref{sec:icw_appendix}. In this appendix, we show that the effect of the weighting in Equation \ref{eq:OQEQuadratic} is to project out the modes of $\textbf{U}$ with a different covariance structure than $\mathbf{S}$ in the power spectrum estimate, and the effect of Equation \ref{eq:OQELinear} is to subtract out the remaining bias. Similar effects for a more realistic model of the EoR and foregrounds are shown in \citet{liu_tegmark2011}.

@@

%We do this when forming the un-normalized power spectrum estimate $\widehat{q}_{\alpha}$:
Motivated by the desire to retain the desirable behavior of suppressing contributions of $\mathbf{U}$ to estimates of the EoR power spectrum, we note that is possible to define a modified version of the quadratic estimator 
%which may still produce an unbiased estimate, 
where Equation \eqref{eq:OQEQuadratic} is replaced by
\begin{equation}
\label{eq:qhat}
\widehat{q}_{\alpha} = \frac{1}{2}\textbf{x}^{\dagger}\textbf{R}\textbf{Q}^{\alpha}\textbf{R}\textbf{x}.
\end{equation}
where $\textbf{R}$ is a weighting matrix chosen by the data analyst.  For example, inverse covariance weighting (the optimal form of QE) would set $\textbf{R} \equiv \textbf{C}^{-1}$ and a uniform-weighted case would use $\textbf{R} \equiv \textbf{I}$, the identity matrix.
The matrix $\textbf{Q}^{\alpha}$ encodes the dependence of the covariance on the power spectrum parameters $\partial \C/\partial p_\alpha$ but in practice also does other things, including implementing a transform of the frequency-domain visibilities to $\mathbf{k}$-space, taking into account cosmological scalings, and converting the visibilities from Jy to Kelvin.
%with dimensions ($N_{f}$,$N_{f}$) --- as

%%% COME BACK HERE
% Need to still explain the motivation for this
With an appropriate normalization matrix $\textbf{M}$, the quantity
%We normalize our power spectrum estimates using the matrix $\textbf{M}$:
\begin{equation}
\label{eq:phat}
\widehat{\textbf{P}} = \textbf{M}\widehat{\textbf{q}}
\end{equation}
is a sensible estimate of the true power spectrum $\textbf{P}$. To ensure that $\textbf{M}$ correctly normalizes our power spectrum, one may take the expectation value of Equation \eqref{eq:phat} to obtain
\begin{eqnarray}
\langle \widehat{P}_\alpha \rangle &=& \frac{1}{2} \sum_{\beta \gamma} M_{\alpha \gamma}  \textrm{tr} \left( \textbf{R}\textbf{Q}^{\gamma}\textbf{R} \textbf{Q}^{\beta}\right) p_\beta +\frac{1}{2} \sum_{ \gamma}  \Tr\left( \mathbf{U} \textbf{R} \textbf{Q}^{\gamma} \mathbf{R} \right) \nonumber \\
&\equiv & \sum_{\beta} W_{\alpha \beta} p_\beta +\frac{1}{2} \sum_{ \gamma}  \Tr\left( \mathbf{U} \textbf{R} \textbf{Q}^{\gamma} \mathbf{R} \right), \label{eq:Wpplusbias}
\end{eqnarray}
where $W_{\alpha \beta}$ are elements of a window function matrix. Consider the first term of this expression. (Again, we are assuming that the linear bias term is significantly suppressed; and if this is not the case, we are simply assuming that we are setting a conservative upper limit). If $\mathbf{W}$ ends up being the identity matrix for our choices of $\mathbf{R}$ and $\mathbf{M}$, then we recover Equation \eqref{eq:super_unbiased} for the first term, and we have an estimator that has no multiplicative matrix bias. However, Equation \eqref{eq:super_unbiased} is a rather restrictive condition, and it is possible to violate it and still have a sensible (and correctly normalized) power spectrum estimates. In particular, as long as the rows of $\mathbf{W}$ sum to unity, our power spectrum will be correctly normalized. Beyond this, the data analyst has a choice for $
\textbf{M}$. For simplicity in this section we choose $\textbf{M}$ to be diagonal, although we explore other cases for the analysis 
of PAPER-64 data as explained in Section \ref{sec:Bias}. In a preview of what is to come, we also stress that the derivation that leads to Equation \eqref{eq:Wpplusbias} assumes that $\mathbf{R}$ and $\mathbf{x}$ are not correlated. If this assumption is violated, a simple application of the (now incorrect) formulae in this section can result in an improperly normalized power spectrum estimator that does not conserve power, i.e., one that has signal loss.
% JEA - is one really free to choose M and R independently?

@@

For the case of a mode which can be safely assumed to be predominantly a foreground, its presence in the covariance matrix will result in the desired suppression via a kind of projection (c.f. Equations \ref{eq:IdealToyModelCovariance} and \ref{eq:IdealToyModelEstimator});  whether or not it is strongly correlated with the the actual data vector is irrelevant.  However, in the case the covariance matrix is empirically estimated, the eigenmodes of $\textbf{C}_{\rm EoR}$ will both be incorrect and correlated with the data.  Incorrect eigenmodes of $\textbf{C}_{\rm EoR}$ that are not correlated with the data will lead to non-minimum variance estimates, but will not produce the suppression of the power spectrum amplitude seen.  As indicated in Appendix \ref{sec:sigloss_appendix}, however, having $\widehat{\mathbf{C}}$ correlated with the data vector $\mathbf{x}$ also produces a kind of projection of power in the {\it non}-foreground modes from the resulting power spectrum estimate, thus producing an estimate that is biased low.  In short, {\it if the covariance is computed from the data itself, it carries the risk of over-fitting information in the data and introducing a multiplicative bias (per $k$) to estimates of the signal.}
